\chapter{Introduction}\label{ch:paper_introduction}

Decentralised Finance (DeFi) is a multi-billion-dollar industry which consists of financial markets and services living on blockchains. Due to the unregulated nature of this space, it is not possible for conventional enforcement mechanisms to ensure parties adhere to deals they've made. Instead, the standard approach is to use programmatic \textbf{smart contract}s which automatically enforce the parameters of whatever deal is being made. Oftentimes, this is not a simple contract between two parties, but a complex self-managing system that can be interacted with by arbitrary users. Examples of such services are decentralized exchanges for converting between currencies, lending platforms for borrowing funds, and many other such applications. Due to the increasing complexity of such protocols, the importance of smart contract security is ever-growing.
\par
The Ethereum blockchain \cite{yellowpaper} is one of the most popular smart contract ecosystems. In this space alone, hundreds of millions of dollars are lost due to smart contract vulnerabilities every year \cite{immunefi} \cite{chainalysis} \cite{slowmist}. Due to the aforementioned unregulated nature and lack of enforcement mechanisms, there is no reliable way of recovering lost funds. As such, tens of millions of dollars are also spent every year on bug bounties, third party audits, and further security measures \cite{immunefi}. Since there is no centralized governing body, much of the ecosystem is based on trust -- most projects are open-source, and users rely on the track records of developers and companies to make informed decisions about investments. Not every user is able to reason about the code or specification of every project, so they instead often rely on stamps of approval given by third-party audit companies. Because automatic tools are only capable of preventing around 8\% of real-world exploits \cite{tool-efficacy}, manual audits dominate this space. Of course, audits are only as reliable as the code is readable -- if a language obscures behaviour through implicit side effects or hidden control flow, even expert auditors can miss vulnerabilities. By requiring all code to be very explicit about what it's doing, as well as avoiding hidden behaviours, we aim to enhance the \textit{auditability} of the language, making audits more thorough and reliable.
\par
The Ethereum blockchain runs the Ethereum Virtual Machine (\textbf{EVM}), which dictates the execution model of smart contracts. Contracts are deployed as immutable bytecode, which is publicly available. However, the source code it was generated from may not be, although the vast majority of public projects publish their code as a way of building trust.
\par
There are two widespread programming languages that are used to generate EVM bytecode. These are \textbf{Solidity} \cite{solidity}, an object-oriented language inspired by a variety of ``curly-braced'' languages, and \textbf{Vyper} \cite{vyper}, which is closely based on Python. Given their similarity with general-purpose programming languages, these languages were not initially designed with built-in safety mechanisms to protect against EVM intricacies. This puts additional burden on developers to preemptively protect against vulnerabilities that are enabled not by their own code, but by the design of the EVM itself.
\par
Designing a language with the peculiarities of the EVM in mind, as well as observing the importance of code auditability, has significant potential to prevent future exploits and reduce the risk involved with deploying new projects (or investing in them). We argue that moving security guarantees into the language design and type system, rather than relying on developer discipline, is the most reliable way of ensuring smart contract safety. Our three core features are motivated by this principle:

\begin{enumerate}
  \item We specifically aim to prevent a vulnerability class called \textbf{reentrancy}, which will be thoroughly explained in chapters \ref{ch:background} and \ref{ch:methodology}.
  \item  In order to improve auditability, we also require the developer to explicitly annotate hidden side effects, incentivizing a more ``effect-oriented'' approach to coding, as well as reducing the probability of unexpected interactions.
  \item We offload responsibility from the developer to the compiler by providing an expressive type system, including \textbf{algebraic data types}.
\end{enumerate}
